---
title: "Data visualisation III  \nMultivariate plots, Maps and Regressions"
author:
- O Rodriguez de Rivera Ortega, PhD\newline
- University of Exeter
#date: "null"

---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)
```

# Grouping

```{r echo=TRUE}
library(curl)
Fires <- read.csv(curl("https://raw.githubusercontent.com/orrortega/statistics-natural-resources/main/data/Largest_fires_by_continent.csv"))
attach(Fires)
```

##
In grouping, the values of the first two variables are mapped to the x and y axes. Then additional variables are mapped to other visual characteristics such as color, shape, size, line type, and transparency. Grouping allows you to plot the data for multiple groups in a single graph.
```{r echo=TRUE,fig.asp=1/2}
library(ggplot2)
ggplot(Fires, 
      aes(x = speed..km.day.1., y=Size..km2.)) + 
      geom_point() +
      labs(title = "Size (km2) by speed (km/day)")
```

##  
Next, let’s include the Regions, using color.
```{r echo=TRUE,fig.asp=2/3}
ggplot(Fires, 
      aes(x = speed..km.day.1., y=Size..km2., 
      color = dominant.spread.direction)) + geom_point() +
      labs(title = "Size (km2) by speed (km/day) and region")
```

##
Finally, let’s add the shape of the trees, using the shape of the points to indicate shape. We’ll increase the point size and add transparency to make the individual points clearer.
```{r echo=TRUE,fig.asp=1/2}
ggplot(Fires, 
      aes(x = speed..km.day.1., y=Size..km2., 
      color = dominant.spread.direction, shape = Region)) + geom_point() +
      labs(title = "Size (km2) by speed (km/day), region and dominant spread direction")
```

##
```{r echo=TRUE,fig.asp=1/2}
ggplot(Fires, 
  aes(x = speed..km.day.1., y=Size..km2., color = Region)) + 
  geom_point(alpha = .4, size = 3) +
  geom_smooth(se=FALSE, method = "lm", formula = y~poly(x,2), size = 1.5) + 
  labs(x = "Speed (km/day)", y = "Size (km2)", title = "Size (km2) by speed (km/day) and region", 
    subtitle = "Summary of Large Forest Fires", y = "", color = "Regions") +
    scale_color_brewer(palette = "Set1") + theme_minimal()
```
Grouping allows you to plot multiple variables in a single graph, using visual characteristics such as color, shape, and size.

##

In **faceting**, a graph consists of several separate plots or small multiples, one for each level of a third variable, or combination of variables. It is easiest to understand this with an example.
```{r echo=TRUE, fig.asp=2/3, message=FALSE}
ggplot(Fires, aes(x = Size..km2.)) + geom_histogram(fill = "cornflowerblue",
    color = "white") + facet_wrap(~Region, ncol = 1) +
    labs(title = "Size (km2) by species")
```
The `facet_wrap` function creates a separate graph for each `species`. The `ncol` option controls the number of columns.

##  
```{r echo=TRUE, fig.asp=1/2, message=FALSE}
ggplot(Fires, aes(x = Size..km2.)) + 
  geom_histogram(fill = "cornflowerblue", color = "white") + 
  facet_wrap(Region ~ dominant.spread.direction) +
  labs(title = "Size histograms by Region")
```
The format of the `facet_grid` function is `facet_grid( row variable(s) ~ column variable(s))`

##

```{r include=FALSE}
library(dplyr)
plotdata <- Fires %>% 
  group_by(Region, year) %>%                            # multiple group columns
  summarise(total = sum(Size..km2.), mean_size = mean(Size..km2.))  # multiple summary columns
```

```{r echo=TRUE,fig.asp=1/2}
# plot total area by year, for each country
ggplot(plotdata, aes(x=year, y = total)) + geom_line(color="grey") + 
geom_point(color="blue") + facet_wrap(~Region) + theme_minimal(base_size = 9) + 
theme(axis.text.x = element_text(angle = 45, hjust = 1)) + 
labs(title = "Changes in burned area",x = "Year",y = "Total area")
```

# Maps

##
```{r echo=TRUE, message=FALSE, warning=FALSE}
library(rworldmap) 
# get map
MyMap <- ggplot() + borders("world", colour="black", fill="grey") + theme_bw()
MyMap
```

```{r}
MyMap  +
  geom_point(data = Fires, aes(x=lon, y=lat),size = 2, color = "red")+
  theme(panel.grid.major = element_line(color = "gray60", linetype = "dashed", size = 0.25), 
        panel.background = element_rect(fill = "aliceblue"))
```

```{r echo=TRUE, message=FALSE, warning=FALSE}
library(viridis)
MyMap  + 
  geom_point(data = Fires, aes(x=lon, y=lat, color = Size..km2.),size = 2) +
  scale_color_viridis() +
  theme(panel.grid.major = element_line(color = "gray60", linetype = "dashed", size = 0.25), 
        panel.background = element_rect(fill = "aliceblue"))
```
## 
```{r echo=TRUE, message=FALSE, warning=FALSE}
library(viridis)
MyMap  + coord_fixed(xlim=c(-10, 37.5), ylim=c(30, 48)) + 
  geom_point(data = Fires, aes(x=lon, y=lat, color = Size..km2.),size = 2) +
  scale_color_viridis() +
  theme(panel.grid.major = element_line(color = "gray60", linetype = "dashed", size = 0.25), 
        panel.background = element_rect(fill = "aliceblue"))
```

## 

```{r echo=TRUE, message=FALSE, warning=FALSE ,fig.asp=1/2}
MyMap  + coord_fixed(xlim=c(-10, 37.5), ylim=c(30, 48)) + 
  geom_point(data = Fires, aes(x=lon, y=lat, color = Size..km2.),size = 2) +
  scale_color_viridis() +
  facet_wrap(~year) +
  theme(panel.grid.major = element_line(color = "gray60", linetype = "dashed", size = 0.25), 
        panel.background = element_rect(fill = "aliceblue"))
```


# Time-dependent graphs

## Time series
```{r include=FALSE}
library(curl)
pm10 <- read.csv(curl("https://raw.githubusercontent.com/orrortega/statistics-natural-resources/main/data/pm10daily.csv"))
pm10$site<-as.factor(pm10$site)
pm10subset <- subset(pm10, site == "Ealing - Hanger Lane Gyratory")
attach(pm10subset)
```

##   
```{r echo=TRUE, message=FALSE, warning=FALSE, fig.asp=2/3}
ggplot(pm10subset, aes(x = day, y = pm10)) +
      geom_line() +
      labs(title = "Levels of pm10 during 2014", x = "Day of the year", y = "pm10 concentration")
```

##  
```{r echo=TRUE, message=FALSE, warning=FALSE, fig.asp=2/3}
library(scales)
ggplot(pm10subset, aes(x = day, y = pm10)) +
    geom_line(color = "indianred3", size=1 ) +
    geom_smooth() +  labs(title = "Pm10 levels", subtitle = "2014", x = "DOY", y = "pm10 concentration") +
    theme_minimal()
```

##  
```{r echo=TRUE, message=FALSE, warning=FALSE, fig.asp=2/3}
# multivariate time series
attach(pm10)
ggplot(pm10,
    aes(x=day , y= pm10, color=site)) + geom_line(size=1) + 
    labs(title = "pm10 levels London",
    subtitle = "2014", caption = "source: OpenAir", y = "pm10 concentration") +
    theme_minimal() + scale_color_brewer(palette = "Dark2")
```

## Area Charts

```{r echo=TRUE, message=FALSE, warning=FALSE, fig.asp=2/3}
ggplot(pm10subset, aes(x = day, y = pm10)) + 
  geom_area(fill="lightblue", color="black") + 
  labs(title = "pm10 levels", x = "day", y = "pm10 concentration")
```

##  
A stacked area chart can be used to show differences between groups over time.

```{r echo=TRUE, message=FALSE, warning=FALSE, fig.asp=2/3}
ggplot(pm10, aes(x = day, y = pm10, fill= site)) + 
  geom_area() + 
  labs(title = "pm10 levels", x = "day", y = "pm10 concentration")
```


# Statistical models

## Correlation plots

The levels of the site variable can be reversed using the fct_rev function in the forcats package.
```{r include=FALSE}
library(curl)
FI_Murcia <- read.csv(curl("https://raw.githubusercontent.com/orrortega/statistics-natural-resources/main/data/FI_Murcia.csv"))
attach(FI_Murcia)
```

```{r echo=TRUE, message=FALSE, warning=FALSE}
df <- dplyr::select_if(FI_Murcia, is.numeric)
r <- cor(df, use="complete.obs") 
round(r,2)
```


##
```{r echo=TRUE, fig.asp=2/3}
library(ggcorrplot) 
ggcorrplot(r)
```
By default, zero count bars are dropped and the remaining bars are made wider. This may not be the behavior you want. You can modify this using the `position = position_dodge(preserve = "single")` option.


##
```{r echo=TRUE, fig.asp=2/3}
ggcorrplot(r,
            hc.order = TRUE,
            type = "lower", lab = TRUE)
```


## Linear Regression
```{r echo=TRUE, fig.asp=1/2}
Height_lm <- lm(Height ~ Diameter1 + Diameter2, data = FI_Murcia)
library(visreg)
#The visreg function takes (1) the model and (2) the variable of interest and plots 
#the conditional relationship, controlling for the other variables. 
#The option gg = TRUE is used to produce a ggplot2 graph.
visreg(Height_lm, "Diameter1", gg = TRUE) +
  labs(title = "Relationship between Height and Diameter1",
  caption = "source:   NFI Murcia",
y = "Height (m)",
x = "Diameter (m)")
```

## Logistic regression
```{r include=FALSE, echo=TRUE, fig.asp=1/2}
Chrod <- read.csv(curl("https://raw.githubusercontent.com/orrortega/statistics-natural-resources/main/data/Chrodatacomplete.csv"))
library(tidyr)
Chrod <- na.omit(Chrod)
attach(Chrod)
Chrod_glm <- glm(presence ~ altitude + Temperature..C. + Rainfall..mm. + Beaufort.scale,
              family="binomial", data=Chrod)
```

```{r echo=TRUE, fig.asp=1/2}
visreg(Chrod_glm, "altitude",
    gg = TRUE,
    scale="response") + labs(y = "Prob(Presence)", x = "Altitude",
    title = "Relationship of Age and Presence")
```

























