---
title: "Data visualisation II  \nUnivariate & Bivariate "
author:
- O Rodriguez de Rivera Ortega, PhD\newline
- University of Exeter
#date: "null"

---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)
```

# Univariate Graphs - Categorical

```{r include=FALSE}
library(curl)
FI_Murcia <- read.csv(curl("https://raw.githubusercontent.com/orrortega/statistics-natural-resources/main/data/FI_Murcia.csv"))
attach(FI_Murcia)
FI_Murcia_sub = subset(FI_Murcia, Species == c("Castanea sativa", "Fagus sylvatica", "Pinus nigra", "Quercus robur"))
```

##   
```{r echo=TRUE,fig.asp=2/3}
library(ggplot2)
# plot the distribution of species
ggplot(FI_Murcia_sub, aes(x = Species)) + 
  geom_bar()
```

##  
```{r echo=TRUE,fig.asp=2/3}
# plot the distribution of race with modified colors and labels
ggplot(FI_Murcia_sub, aes(x = Species)) + 
  geom_bar(fill = "cornflowerblue", color="black") + 
  labs(x = "Species", y = "Frequency", 
       title = "Individuals by species")
```

##  
```{r echo=TRUE,fig.asp=2/3}
# plot the distribution as percentages
ggplot(FI_Murcia_sub, aes(x = Species, y = ..count.. / sum(..count..))) + 
  geom_bar() + labs(x = "Species", y = "Percent", 
                    title = "Individuals by species") + 
  scale_y_continuous(labels = scales::percent)
```

##     
```{r echo=TRUE,fig.asp=2/3}
# calculate number of participants in # each race category
library(dplyr)
plotdata <- FI_Murcia_sub %>%
  count(Species)
plotdata
```

##  
```{r echo=TRUE,fig.asp=2/3}
# plot the bars in ascending order
ggplot(plotdata, aes(x = reorder(Species, n), y = n)) + 
  geom_bar(stat = "identity") +
  labs(x = "Species", y = "Frequency", 
       title = "Individuals by species")
```

##  
```{r echo=TRUE,fig.asp=2/3}
# plot the bars in descending order
ggplot(plotdata, aes(x = reorder(Species, -n), y = n)) + 
  geom_bar(stat = "identity") +
  labs(x = "Species", y = "Frequency", 
       title = "Individuals by species")
```

##   
```{r echo=TRUE,fig.asp=1/2}
# plot the bars with numeric labels
# geom_text adds the labels, 
# vjust controls vertical justification
ggplot(plotdata, aes(x = reorder(Species, n), y = n)) + 
  geom_bar(stat = "identity") + 
  geom_text(aes(label = n), vjust=-0.5) +
  labs(x = "Species", y = "Frequency", 
       title = "Individuals by species")
```
##  
```{r}
plotdata <- FI_Murcia_sub %>% count(Species) %>% mutate(pct = n / sum(n),
pctlabel = paste0(round(pct*100), "%"))
```

```{r}
# plot the bars as percentages,
# in descending order with bar labels 
ggplot(plotdata, aes(x = reorder(Species, -pct), y = pct)) +
  geom_bar(stat = "identity", fill = "indianred3", color = "black") + 
  geom_text(aes(label = pctlabel), vjust = -0.25) + 
  scale_y_continuous(labels = scales::percent) +
  labs(x = "Species", y = "Percent", title = "Individuals by species")
```
## Overlapping labels
```{r}
# horizontal bar chart
ggplot(FI_Murcia_sub, aes(x = Species)) + 
  geom_bar() +
  labs(x = "", y = "Frequency", title = "Individuals by shape") + coord_flip()
```

# Univariate Graphs - Quantitative

## Histogram
```{r echo=TRUE, message=FALSE, warning=FALSE ,fig.asp=1/2}
# plot the age distribution using a histogram
ggplot(FI_Murcia_sub, aes(x = Height)) + geom_histogram() +
  labs(title = "Individuals by Height", x = "Height")
```

##   
```{r echo=TRUE, message=FALSE, warning=FALSE}
# plot the histogram with blue bars and white borders
ggplot(FI_Murcia_sub, aes(x = Height)) + 
  geom_histogram(fill = "cornflowerblue", color = "white") + 
  labs(title = "Individuals by Height", x = "Height")
```

##  
```{r echo=TRUE, message=FALSE, warning=FALSE}
# plot the histogram with 20 bins
ggplot(FI_Murcia_sub, aes(x = Height)) + 
  geom_histogram(fill = "cornflowerblue", color = "white", bins = 20) + 
  labs(title="Individuals by Height", subtitle = "number of bins = 20", x = "Height")
```

##  
```{r echo=TRUE, message=FALSE, warning=FALSE}
# plot the histogram with a binwidth of 5
ggplot(FI_Murcia_sub, aes(x = Height)) + 
  geom_histogram(fill = "cornflowerblue", color = "white", binwidth = 5) + 
  labs(title="Individuals by Height", subtitle = "binwidth = 5 meters", x = "Height")
```

##  
```{r echo=TRUE, message=FALSE, warning=FALSE}
# plot the histogram with percentages on the y-axis
library(scales) 
ggplot(FI_Murcia_sub, aes(x = Height, y= ..count.. / sum(..count..))) +
  geom_histogram(fill = "cornflowerblue", color = "white", binwidth = 5) + 
  labs(title="Individuals by Height", y = "Percent", x = "Height") + 
  scale_y_continuous(labels = percent)
```

##  
There are more things to add some "fantasy" to your plots as:

* Kernel density plot
* Dotplot
* etc.

# Bivariate Graphs

## Categorical vs Categorical

##  
```{r echo=TRUE, fig.asp=1/2}
# stacked bar chart
FI_Murcia_sub$Shape<- as.factor(FI_Murcia_sub$Shape)
ggplot(FI_Murcia_sub,
aes(x = Species,
fill = Shape)) + geom_bar(position = "stack")
```
From the chart, we can see for example, that the most common species is F. sylvatica. An the most common shape for all the species is 2

**Stacked is the default, so the last line could have also been written as `geom_bar()`.**

##  Grouped bar chart
```{r echo=TRUE, fig.asp=1/3}
ggplot(FI_Murcia_sub,
      aes(x = Species,
          fill = Shape)) + geom_bar(position = "dodge")
```
By default, zero count bars are dropped and the remaining bars are made wider. This may not be the behavior you want. You can modify this using the `position = position_dodge(preserve = "single")` option.


##
```{r echo=TRUE, fig.asp=2/3}
ggplot(FI_Murcia_sub,
      aes(x = Species,
          fill = Shape)) + 
        geom_bar(position = position_dodge(preserve="single"))
```


##
```{r echo=TRUE, fig.asp=2/3}
#Segmented bar chart
FI_Murcia_sub$Shape<- as.factor(FI_Murcia_sub$Shape)
ggplot(FI_Murcia_sub,
      aes(x = Species,
          fill = Shape)) + 
        geom_bar(position = "fill") + labs(y="Proportion")
```
This type of plot is particularly useful if the goal is to compare the percentage of a category in one variable across each level of another variable.

## Quantitative vs Quantitative

##
The simplest display of two quantitative variables is a scatterplot, with each variable represented on an axis.
```{r echo=TRUE, fig.asp=1/2}
# simple scatterplot
ggplot(FI_Murcia_sub,
aes(x = Diameter1, y = Height)) + 
  geom_point()
```

##
```{r echo=TRUE, fig.asp=1/2}
# scatterplot with linear fit line
ggplot(FI_Murcia_sub,
aes(x = Diameter1, y = Height)) + 
  geom_point(color="steelblue") +
  geom_smooth(method = "lm")
```
Clearly, Height increases with Diameter. However a straight line does not capture this non-linear effect. A polynomial regression line will fit better here.

##
```{r echo=TRUE, fig.asp=1/2}
# scatterplot with quadratic line of best fit
ggplot(FI_Murcia_sub,
aes(x = Diameter1, y = Height)) + 
  geom_point(color="steelblue") +
  geom_smooth(method = "lm", formula = y ~ poly(x, 2),
  color = "indianred3")
```

##
Finally, a smoothed nonparametric fit line can often provide a good picture of the relationship. The default in **`ggplot2`** is a loess line which stands for for locally weighted scatterplot smoothing.

```{r echo=TRUE, fig.asp=1/2}
# scatterplot with quadratic line of best fit
ggplot(FI_Murcia_sub,
aes(x = Diameter1, y = Height)) + 
  geom_point(color = "steelblue") +
  geom_smooth(color = "tomato")
```

## Line plot
When one of the two variables represents time, a line plot can be an effective method of displaying relationship.
```{r echo=TRUE, fig.asp=1/2}
library(curl)
rats <- read.csv(curl("https://raw.githubusercontent.com/orrortega/statistics-natural-resources/main/data/LeslieandDaviesRats.csv"))
ggplot(rats, 
       aes(x = time, y = captures)) + 
  geom_line()
```

##
```{r echo=TRUE, fig.asp=1/2}
# line plot with points and improved labeling
ggplot(rats,
  aes(x = time, y = captures)) +
geom_line(size = 1.5, color = "lightgrey") + geom_point(size = 3, color = "steelblue") +
labs(y = "Individuals captured", x = "Ocasion",
title = "Removal method data", subtitle = "An Attempt to Determine the Absolute Number of Rats on a Given Area",
caption = "Leslie and Davies (1939)")
```
**Time dependent data is covered in more detail under Time series.**

## Categorical vs Quantitative

When plotting the relationship between a categorical variable and a quantitative variable, a large number of graph types are available.

These include bar charts using summary statistics, grouped kernel density plots, side-by-side box plots, side-by-side violin plots, and Cleveland plots.

## Bar chart (on summary statistics)
```{r echo=TRUE, fig.asp=1/2, message=FALSE}
# calculate mean salary for each rank
plotdata <- FI_Murcia_sub %>%
    group_by(Species) %>% 
    summarize(mean_Height = mean(Height))
# plot mean salaries
ggplot(plotdata, 
       aes(x = Species, y = mean_Height)) + 
  geom_bar(stat = "identity")
```

```{r echo=TRUE, fig.asp=2/3}
# plot mean heights in a more attractive fashion
library(scales) 
ggplot(plotdata,
      aes(x = factor(Species,
          labels = c("Castanea\nsativa", "Fagus\nsylvatica", "Pinus\nnigra", "Quercus\nrobur")), y = mean_Height)) +
    geom_bar(stat = "identity", fill = "cornflowerblue") +
    geom_text(aes(label = round(mean_Height, 2)), 
      vjust = -0.2) +
  labs(title = "Mean Height by tree species",
    subtitle = "National Forest Inventory (Murcia)", x = "",
    y = "m")
```

## Grouped kernel density plots
One can compare groups on a numeric variable by superimposing kernel density plots in a single graph.
```{r echo=TRUE, fig.asp=1/2}
# by rank using kernel density plots 
ggplot(FI_Murcia_sub,
    aes(x = Height, fill = Species)) +
  geom_density(alpha = 0.4) +
  labs(title = "Height distribution by species")
```
`alpha` values range from 0 (transparent) to 1 (opaque)

## Box plots

```{r echo=TRUE, fig.asp=1/2}
# plot the distribution of salaries by rank using boxplots
ggplot(FI_Murcia_sub,
      aes(x = Species, y = Height)) + 
  geom_boxplot() +
labs(title = "Height distribution by species")
```

## Violin plots
Violin plots are similar to kernel density plots, but are mirrored and rotated 90$^\circ$ .
```{r echo=TRUE, fig.asp=1/2}
 # plot the distribution of salaries by rank using violin plots 
ggplot(FI_Murcia_sub,
      aes(x = Species, y = Height)) +
      geom_violin() +
      labs(title = "Height distribution by species")
```

##
A useful variation is to superimpose boxplots on violin plots.
```{r echo=TRUE, fig.asp=1/2}
# plot the distribution of salaries by rank using violin plots 
ggplot(FI_Murcia_sub,
      aes(x = Species, y = Height)) +
      geom_violin(fill = "cornflowerblue") + 
      geom_boxplot(width = .2, fill = "orange", outlier.color = "orange", outlier.size = 2) +
      labs(title = "Height distribution by species")
```

## Cleveland Dot Charts
Cleveland plots are useful when you want to compare a numeric statistic for a large number of groups.
```{r echo=TRUE, fig.asp=2/3}
# plot the distribution of salaries by rank using violin plots 
plotdata<-FI_Murcia %>% 
    group_by(Species) %>%
    mutate(Mean = mean(Height, na.rm=TRUE))
ggplot(plotdata,
      aes(x = Mean, y = Species)) +
      geom_point()
```

##
```{r echo=TRUE, fig.asp=2/3}
# plot the distribution of salaries by rank using violin plots 
ggplot(plotdata,
      aes(x = Mean, y = reorder(Species,Mean))) +
      geom_point()
```