---
title: "Discriminant Analysis"
author:
- O Rodriguez de Rivera Ortega, PhD\newline
- University of Exeter
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Introduction

linear discriminant analysis, originally developed by R A Fisher in 1936 to classify subjects into one of the two clearly defined groups.

It was later expanded to classify subjects into more than two groups.

Linear Discriminant Analysis (LDA) is a dimensionality reduction technique. LDA used for dimensionality reduction to reduce the number of dimensions (i.e. variables) in a dataset while retaining as much information as possible.

Basically, it helps to find the linear combination of original variables that provide the best possible separation between the groups.

The basic purpose is to estimate the relationship between a single categorical dependent variable and a set of quantitative independent variables.

The major applications or examples are

Predicting success or failure of new products
Accepting or rejecting admission to an applicant.
Predicting credit risk category for a person
Classifying patients into different categories.


```{r}
library(car)
#install.packages('rattle')
library(rattle)
data(wine, package='rattle')
attach(wine)
head(wine)
```



```{r}
scatterplotMatrix(wine[2:6])
```

The purpose of linear discriminant analysis (LDA) in this example is to find the linear combinations of the original variables (the 13 chemical concentrations here) that gives the best possible separation between the groups (wine cultivars here) in our data set. Linear discriminant analysis is also known as “canonical discriminant analysis”, or simply “discriminant analysis”.

If we want to separate the wines by cultivar, the wines come from three different cultivars, so the number of groups G=3, and the number of variables is 13 (13 chemicals’ concentrations; p=13). The maximum number of useful discriminant functions that can separate the wines by cultivar is the minimum of G−1 and p, and so in this case it is the minimum of 2 and 13, which is 2. Thus, we can find at most 2 useful discriminant functions to separate the wines by cultivar, using the 13 chemical concentration variables.

You can carry out a linear discriminant analysis using the “lda()” function from the R MASS package.

```{r}
# install.packages('MASS')
library(MASS)
wine.lda <- lda(Type ~ ., data=wine)
```

To get the values of the loadings of the discriminant functions for the wine data, we can type:

```{r}
wine.lda
```

This means that the first discriminant function is a linear combination of the variables: −0.403∗Alcohol+0.165∗Malic⋯−0.003∗Proline. For convenience, the value for each discriminant function (eg. the first discriminant function) are scaled so that their mean value is zero and its variance is one.

The “proportion of trace” that is printed when you type “wine.lda” (the variable returned by the lda() function) is the percentage separation achieved by each discriminant function. For example, for the wine data we get the same values as just calculated (68.75% and 31.25%).

A Stacked Histogram of the LDA Values
A nice way of displaying the results of a linear discriminant analysis (LDA) is to make a stacked histogram of the values of the discriminant function for the samples from different groups (different wine cultivars in our example).

We can do this using the “ldahist()” function in R. For example, to make a stacked histogram of the first discriminant function’s values for wine samples of the three different wine cultivars, we type:

```{r}
wine.lda.values <- predict(wine.lda)
ldahist(data = wine.lda.values$x[,1], g=Type)
```

second discriminant function separates those cultivars, by making a stacked histogram of the second discriminant function’s values:

```{r}
ldahist(data = wine.lda.values$x[,2], g=Type)
```

## Scatterplots of the Discriminant Functions

We can obtain a scatterplot of the best two discriminant functions, with the data points labelled by cultivar, by typing:

```{r}
plot(wine.lda.values$x[,1],wine.lda.values$x[,2]) # make a scatterplot
text(wine.lda.values$x[,1],wine.lda.values$x[,2],Type,cex=0.7,pos=4,col="red") # add labels
```

discriminant functions, we can see that the wines from the three cultivars are well separated in the scatterplot. The first discriminant function (x-axis) separates cultivars 1 and 3 very well, but doesn’t not perfectly separate cultivars 1 and 3, or cultivars 2 and 3.

The second discriminant function (y-axis) achieves a fairly good separation of cultivars 1 and 3, and cultivars 2 and 3, although it is not totally perfect.

To achieve a very good separation of the three cultivars, it would be best to use both the first and second discriminant functions together, since the first discriminant function can separate cultivars 1 and 3 very well, and the second discriminant function can separate cultivars 1 and 2, and cultivars 2 and 3, reasonably well. 