---
title: "Extended Time Series Analysis in R"
author: "O. Rodriguez de Rivera Ortega, PhD | University of Exeter"
date: "2025-05-21"
output: html_document
editor_options: 
  markdown: 
    wrap: 72
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(forecast)
library(TTR)
library(datasets)
```

# Introduction to Time Series

cat("A time series is a set of observations recorded sequentially over
time, typically at regular intervals.\n") cat("Time series analysis aims
to understand underlying structure and forecast future values.\n")
cat("Key components include:\n") cat("- Trend: long-term movement\n")
cat("- Seasonality: systematic variation at fixed periods\n") cat("-
Noise: random irregular variation\n")

# Moving Average Example

cat("\nWe apply a 3-point moving average to rainfall data to smooth
fluctuations and observe the trend.\n")

```{r}
rain <- scan("http://robjhyndman.com/tsdldata/hurst/precip1.dat", skip=1)
rainseries <- ts(rain, start=c(1813))
rainseriesSMA3 <- SMA(rainseries, n=3)
plot.ts(rainseriesSMA3, main="3-Point Moving Average")

cat("Output: The plot displays a smoothed version of the rainfall time series, which helps highlight long-term trends.\n")
```

# Autocorrelation and Partial Autocorrelation

cat("\nAutocorrelation measures correlation with lagged versions of
itself.\n") cat("Partial autocorrelation shows the direct correlation at
a given lag after controlling for intermediate lags.\n")

```{r}
acf(rainseries, lag.max=20, main="ACF of Rainfall Series")
pacf(rainseries, lag.max=20, main="PACF of Rainfall Series")

cat("Output: Spikes outside the blue bounds indicate statistically significant autocorrelations. Useful for ARIMA modeling.\n")
```

# Seasonal Time Series: CO2

cat("\nWe use the built-in CO2 dataset, which records monthly
atmospheric CO₂ concentrations.\n")

```{r}
ppmtimeseries <- ts(co2, frequency=12, start=c(1959))
plot.ts(ppmtimeseries, main="Monthly CO2 Levels at Mauna Loa")

cat("Output: This time series shows a clear upward trend and recurring seasonal fluctuations.\n")
```

# Decomposition of CO2

cat("\nDecomposition separates the time series into trend, seasonal, and
residual components.\n")

```{r}
ppmtimeseriescomponents <- decompose(ppmtimeseries)
plot(ppmtimeseriescomponents)

cat("Output: The trend component shows long-term growth, seasonal shows repetitive yearly changes, and the remainder is noise.\n")
```

# Seasonal Adjustment

cat("\nSeasonal adjustment removes the seasonal component to better
analyze the underlying trend and irregularities.\n")

```{r}
ppmtimeseriesadjusted <- ppmtimeseries - ppmtimeseriescomponents$seasonal
plot(ppmtimeseriesadjusted, main="Seasonally Adjusted CO₂ Series")

cat("Output: This adjusted series highlights the trend and irregular movements without seasonal interference.\n")
```

# Exponential Smoothing

cat("\nExponential smoothing applies weights that decrease exponentially
to past data.\n") cat("We use Holt-Winters smoothing without trend or
seasonal components.\n")

```{r}
rainseriesforecast <- HoltWinters(rainseries, beta=FALSE, gamma=FALSE)
plot(rainseriesforecast)

cat("Output: The fitted line smooths the rainfall series by responding more to recent values.\n")
```

# Forecasting

cat("\nWe forecast 8 future values using the exponential smoothing
model.\n")

```{r}
rainforecast <- forecast(rainseriesforecast, h=8)
plot(rainforecast)

cat("Output: Forecasts are shown with 80% and 95% prediction intervals.\n")
```

# Residual Diagnostics

cat("\nWe examine residuals to check if the model captured all
information.\n")

```{r}
# Residual diagnostics (handle NAs safely)
clean_residuals <- na.omit(rainforecast$residuals)

acf(clean_residuals, lag.max = 20)
Box.test(clean_residuals, lag = 20, type = "Ljung-Box")
plot.ts(clean_residuals, main = "Residuals (NA removed)")

cat("Output: If residuals show no significant autocorrelation (ACF within bounds, high p-value), the model is adequate.\n")
```

# Forecast Error Distribution

cat("\nWe compare residual distribution to a normal distribution.\n")

```{r}
plotForecastErrors <- function(forecasterrors) {
  mybinsize <- IQR(forecasterrors, na.rm=TRUE)/4
  mysd <- sd(forecasterrors, na.rm=TRUE)
  mymin <- min(forecasterrors, na.rm=TRUE) - mysd*5
  mymax <- max(forecasterrors, na.rm=TRUE) + mysd*3
  mynorm <- rnorm(10000, mean=0, sd=mysd)
  mybins <- seq(min(mymin, min(mynorm)), max(mymax, max(mynorm)), mybinsize)
  hist(forecasterrors, col="red", freq=FALSE, breaks=mybins)
  myhist <- hist(mynorm, plot=FALSE, breaks=mybins)
  points(myhist$mids, myhist$density, type="l", col="blue", lwd=2)
}
plotForecastErrors(rainforecast$residuals)

cat("Output: The red histogram (residuals) should closely match the blue normal curve if residuals are normally distributed.\n")
```

# ARIMA Example: Volcanic Dust Index

cat("\nWe analyze the annual Volcanic Dust Index from 1500 onward.\n")

```{r}
volcanodust <- scan("http://robjhyndman.com/tsdldata/annual/dvi.dat", skip=1)
volcanodustseries <- ts(volcanodust, start=c(1500))
plot.ts(volcanodustseries, main="Volcanic Dust Index")

cat("Output: The plot shows irregular spikes due to major volcanic events. The series is fairly stationary.\n")
```

# ACF/PACF for ARIMA

cat("\nWe use ACF and PACF to identify ARIMA structure.\n")

```{r}
acf(volcanodustseries, lag.max=20)
pacf(volcanodustseries, lag.max=20)

cat("Output: Significant lags help determine AR and MA orders. For example, 2 significant lags in PACF suggests AR(2).\n")
```

# Fit ARIMA and Forecast

cat("\nWe fit an ARIMA(2,0,0) model and forecast 31 future values.\n")

```{r}
volcano_arima <- arima(volcanodustseries, order=c(2,0,0))
volcano_forecast <- forecast(volcano_arima, h=31)
plot(volcano_forecast)

cat("Output: The forecast shows expected dust levels and confidence intervals for the next 31 years.\n")
```

# ARIMA Residual Check

cat("\nWe check residuals from the ARIMA model to assess goodness of
fit.\n")

```{r}
acf(volcano_forecast$residuals, lag.max=20)
Box.test(volcano_forecast$residuals, lag=20, type="Ljung-Box")
plotForecastErrors(volcano_forecast$residuals)

cat("Output: Ideally, no autocorrelation and residuals are normally distributed. If so, ARIMA fit is adequate.\n")
```


